Neighborhoods and Felony Disenfranchisement: The Case of New York City

### 2019-06-06 markdown/kmorris_thesis.Rmd
continue incorporating IV

---
title: "Neighborhoods and Felony Disenfranchisement: The Case of New York City"
author: "Kevin Morris"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
bibliography: "../research/citations.bib"
link-citations: yes
---
<!-- --- -->
<!-- title: "Neighborhoods and Felony Disenfranchisement: The Case of New York City" -->
<!-- author: "Kevin Morris" -->
<!-- date: "`r format(Sys.time(), '%B %d, %Y')`" -->
<!-- output: word_document -->
<!-- bibliography: "../research/citations.bib" -->
<!-- link-citations: yes -->
<!-- always_allow_html: yes -->
<!-- --- -->

```{r setup, echo = F, message = F, warning = F}
knitr::opts_knit$set(root.dir = "..", dev = "cairo_pdf")
library(sandwich)
library(lmtest)
library(scales)
library(tidyverse)
library(extrafont)
library(kableExtra)
library(stargazer)
loadfonts()
```
# Introduction {-}

@@

## Identifying Parolees Whose Rights Were Restored {-}
Following Executive Order 181, the Department of Corrections and Community Supervision began indicating on their online Parolee Lookup Tool whether a parolee had her voting rights restored.^[Only individuals sentenced to prison for felony offenses lose their voting rights in New York. As such, I remove individuals sentenced to prison for misdemeanors. Incarceration likely also has a *de facto* disenfranchising effect on turnout among individuals incarcerated for misdemeanors, though exploring that effect is beyond the scope of this project.] By using the DIN number provided from the parolee public records request, I was able to identify individuals who did and did not have their voting rights restored.^[Not all parolees listed in the public records request data are included in the lookup tool. For individuals who finished parole between January 1^st^, 2018, and April 17^th^, 2018, 1.0 percent are not in the lookup tool. For those discharged from parole between April 18^th^, 2018, and January 13^th^, 2019 (the latest date of the parole records), 1.2 percent of individuals are not found in the lookup tool.]

Although Governor Cuomo signed the Executive Order in April of 2018, an examination of the individuals whose rights were ultimately restored indicates that the program may not have gone into full effect until later in May. As Figure \ref{fig:restoration-chart} demonstrates, individuals who finished their parole supervision in early May did not have their voting rights formally restored until they finished their term of supervision. It is not until May 21^st^ that the majority of individuals being discharged from parole had their voting rights restored prior to their discharge. However, even after the policy change went into effect, not all parolees had their voting rights restored prior to being discharged from parole. These are likely individuals who are not citizens, and therefore ineligible to register to vote.
```{r restoration-chart-chunk, echo = FALSE, message = F, warning = F, fig.width = 7.5, fig.height = 4, fig.cap = "\\label{fig:restoration-chart}Share of Discharged Parolees Whose Voting Rights Were Restored Prior to Discharge", fig.pos = "H", fig.align = "center"}
j <- readRDS("./temp/restoration_plot.rds")
print(j)
```
Figure \ref{fig:restoration-chart} complicates our ability to cleanly identify the impact of the policy change. If no parolees discharged prior to May 21^st^ had had their rights restored, but all parolees discharged on or after May 21^st^ had, identifying those impacted by the policy would be straightforward. That is not the case: some select individuals seem to have had their rights restored before the policy change was fully implemented. Similarly, we would ideally exclude all individuals who did not have their rights restored after the policy change as well as the individuals who were discharged before the change who would not have had their rights restored if they had been discharged from parole later. However, identifying individuals whose rights would not have been restored is not possible.

In the analyses that follow, I consider all parolees discharged on or after May 21^st^, 2018, beneficiaries of the policy change, whether their rights were restored or not. I assume that the share of individuals discharged after May 21^st^ whose rights were not restored is comparable to the share of individuals discharged prior to May 21^st^ whose rights would not have been restored if they had been discharged later. I also remove from the analyses that follow the `r readRDS("./temp/count_restored_before_may21.rds")` individuals discharged from parole prior to May 21^st^, 2018, whose rights were restored prior to their discharge.

## Trends in Turnout {-}
Before analyzing turnout in the 2018 midterms, I begin by examining turnout in the 2016 election. It is possible that individuals discharged from parole shortly before a federal election are more likely to cast a ballot than individuals discharged earlier. If individuals who were discharged from parole prior to May 21^st^, 2016, were less likely to vote than those discharged between May 21^st^ and Election Day, disentangling the impact of the policy change in 2018 from the expected uptick would be more complicated. However, as Figure \ref{fig:to-16} makes clear, individuals discharged from parole in the final months before the 2016 presidential election were not substantially more likely to cast a ballot in the election than individuals discharged earlier in the year. The longer that individuals have been off of parole, the more likely they are to cast a ballot. For instance, of the individuals last discharged from parole in 2010, `r percent(readRDS("./temp/turnout_in_16_finished_10.rds"), accuracy = 0.1)` cast a ballot in the 2016 election, while just `r percent(readRDS("./temp/turnout_in_16_finished_15.rds"), accuracy = 0.1)` of those last discharged from parole in 2015 did so.^[Figure \ref{fig:to-16} plots individuals' turnout by the last date of discharge from parole. Therefore, individuals discharged from parole in 2010 who reoffended and were discharged from parole again in 2015 are included in the denominator only in 2015.] Figure \ref{fig:to-16} plots turnout rates by month of parole discharge. A quadratic curve is fitted (weighted by the number of individuals discharged each month), along with a 95 percent confidence band. This curve is fit on monthly data running from January, 2010 through May, 2016, and extended through October, 2016.

```{r to-16-chart, echo = FALSE, message = F, warning = F, fig.width = 7.5, fig.height = 4, fig.cap = "\\label{fig:to-16}Turnout in 2016 Presidential Election", fig.pos = "H", fig.align = "center"}
j <- readRDS("./temp/to16_chart.rds")
print(j)
```
Figure \ref{fig:to-18} plots month of parole discharge and turnout in the 2018 midterm elections. Once again, a weighted quadratic curve is fitted with a 95 percent confidence band. This curve is fit on monthly data running from January, 2012 through May, 2018, and extended through October, 2018.
```{r to-18-chart, echo = FALSE, message = F, warning = F, fig.width = 7.5, fig.height = 4, fig.cap = "\\label{fig:to-18}Turnout in 2018 Midterm Election", fig.pos = "H", fig.align = "center"}
j <- readRDS("./temp/to18_chart.rds")
print(j)
```
Figure \ref{fig:to-16} does not indicate that individuals who were discharged from parole shortly before the 2016 presidential election were more likely to cast a ballot than individuals discharged earlier in the year. Figure \ref{fig:to-18}, on the other hand, indicates that New Yorkers discharged from parole in the months leading up to the 2018 election --- New Yorkers whose rights were restored while they were still on parole --- were more likely to participate than those discharged earlier in the year. However, Figures \ref{fig:to-16} and \ref{fig:to-18} are noisy. In the section that follows, I develop a logit model to explore whether, after controlling for the available characteristics, individuals discharged from parole after May 21^st^, 2018, were more likely to cast a ballot in the 2018 midterm elections.

@@

Figure \ref{fig:to-18} displays the share of individuals who were discharged from parole in each month and cast a ballot in the 2018 elections. It does not control for any sort of individual level characterisitcs. In \ref{tab:to-18-logit}, I present the results of an individual-level regression exploring the impact of Executive Order 181 on turnout in 2018. Because turnout is a binary dependent variable, this is a logistic regression. It includes all individuals discharged from parole between January 1^st^, 2012, through October 10^th^, 2018 (the registration deadline in New York State).^[This of course ignores the individuals who were still on parole and cast a ballot in the 2018 election. However, for this exercise, I limit the analysis group to individuals who would have been eligible to vote on election day even if Governor Cuomo had not signed Executive Order 181.]

@@

stargazer(model1, model2, model3,
          header = F,
          type = "latex",
          covariate.labels = c("D(Discharged on or after May 21\\textsuperscript{st}, 2018)", "Days Since Discharged",
                               "Days Since Discharged\\textsuperscript{2}", "D(Male)", "Age (Years)", "Counts in Most Recent Sentence",
                               "Time Spent on Parole (Years)"),
          dep.var.labels = c("Cast Ballot in 2018 Election"),
          title = "(\\#tab:to-18-logit) Individual-Level Logit Model",
          table.placement = "H",
          omit = c("county", "race", "felony"),
          add.lines=list(c("County Fixed Effects", "", "X", "X"),
                         c("Race / Ethnicity Fixed Effects", "", "X", "X"),
                         c("Felony Class Fixed Effects", "", "", "X")),
          table.layout = "-dm#-t-a-s-n")
```

Model 1 in Table \ref{tab:to-18-logit} formalizes the trend presented in Figure \ref{fig:to-18} by controlling only for whether an individual was discharged on or after May 21^st^, 2018, and the number of days between the individual's discharge date and November 6^th^, 2018. Model 2 also controls for individual-level characteristics: sex, age on November 6^th^, 2018, county, and race. Model 3 adds sentence-specific information to Model 2: the number of counts in the individual's most recent sentence, the amount of time they spent on parole, and the class(es) of felony for which they were convicted. In each successive model, the AIC decreases substantially, indicating that including these controls are warranted. Table \ref{tab:to-18-logit} makes clear that formerly incarcerated men were far less likely to vote than formerly incarcerated women; that older formerly incarcerated individuals were more likely to cast a ballot; and individuals who spent longer on parole were more likely to participate in the midterm election.

Each model also indicates that Executive Order 181 was successful at increasing turnout among formerly incarcerated New Yorkers.^[These models do not explicitly control for the potentiality that individuals discharged from parole shortly before an election always turnout at higher rates. In [Appendix C], I run the models from Table \ref{tab:to-18-logit} on turnout in the 2016 presidential election. [Appendix C] shows that in 2016, individuals discharged from parole in the months leading up to the election were no more likely to participate in that election] Exponentiating the coefficients on D(Discharged on or after May 21^st^, 2018) indicates that individuals who were discharged after Executive Order 181 went into effect were between `r round((exp(model1[["coefficients"]][["finished_postTRUE"]]) - 1) * 100, digits = 1)` and `r round((exp(model3[["coefficients"]][["finished_postTRUE"]]) - 1) * 100, digits = 1)` percent more likely to vote.

Of course, in these models we are *not* directly testing the impact of having ones rights restored on propensity to vote. In the next section, I employ an instrumental variables approach to confirm that Executive Order was in fact successful at raising turnout.

@@

In the previous section, I explored whether individuals who were discharged from parole after Executive Order 181 went into effect were more likely to vote, whether or not their voting rights were actually restored. There is reason to believe that whether an individual received the treatment (had their rights restored) is correlated with their propensity to vote. For instance, parolees who were arrested did not have their rights restored; a recent arrest is also likely to impact turnout. Similarly, noncitizens did not have voting rights restored / granted; such citizenship status is also correlated with propensity to vote. Because the treatment was not randomly assigned, direct comparison of turnout rates among individuals who did and did not have their rights restored is not probitive to the effect of Executive Order 181. This calls for an instrumental variables approach.

Although \ref{fig:to-18} shows that there is a relationship between how long an individual has been off parole and their propensity to vote over the long run, this relationship is far less signficant in the short run. Figure \ref{fig:to-18-recent} shows that, for individuals discharged from parole in 2017 and 2018, there does not appear to be a relationship between time off parole and turnout rates.

@@

Formalizing this chart into an individual-level logit model demonstrates that this relationship does not hold over the short run. The models in Table \ref{tab:to-18-logit-short} include individuals last discharged from parole between January 1^st^, 2017, and May 20^th^, 2018.

@@

Adding controls for time in Table \ref{tab:to-18-logit-short} increases the AIC, indicating that the inclusion of these variables is not warranted. A Chi-squared test confirms that the model is not improved when controls for time are included.

Although date of discharge is not correlated with turnout over the 2017 -- 2018 period, it is clear from Figure \ref{fig:restoration-chart} that the date of discharge from parole *is* correlated with formal rights restoration. Although not all individuals released from parole after May 21^st^ had their rights formally restored, many did. A dummy variable indicating whether an individual was released on or after May 21^st^, 2018, therefore, is a good instrument for formal rights restoration.

@@

Table \ref{tab:to-18-logit} shows that individuals who were discharged from parole after Governor Cuomo signed Executive Order 181 were substantially more likely to cast a ballot in the 2018 election than individuals who were discharged prior to the policy change. It is, of course, possible that individuals discharged from parole immediately before an election are always more likely to vote, due perhaps to the recency of their interation with state systems and a desire to have their voices heard. In that case, claiming that Executive Order 181 caused turnout to increase would be a mistake.

To test whether this is true --- that individuals discharged shortly before an election turn out at higher rates --- I fit the models from Table \ref{tab:to-18-logit} on 2016 turnout. Table \ref{tab:to-16-logit} includes all individuals last discharged from parole between January 1^st^, 2010, and October 14^th^, 2010.
```{r logit-reg-chunk-16, echo = F, results = "asis", message = F}
load("./temp/individual_turnout_16.rdata")

stargazer(model1, model2, model3,
          header = F,
          type = "latex",
          covariate.labels = c("D(Discharged on or after May 21\\textsuperscript{st}, 2016)", "Days Since Discharged",
                               "Days Since Discharged\\textsuperscript{2}", "D(Male)", "Age (Years)", "Counts in Most Recent Sentence",
                               "Time Spent on Parole (Years)"),
          dep.var.labels = c("Cast Ballot in 2016 Election"),
          title = "(\\#tab:to-16-logit) Individual-Level Logit Model",
          table.placement = "H",
          omit = c("county", "race", "felony"),
          add.lines=list(c("County Fixed Effects", "", "X", "X"),
                         c("Race / Ethnicity Fixed Effects", "", "X", "X"),
                         c("Felony Class Fixed Effects", "", "", "X")),
          table.layout = "-dm#-t-a-s-n")
```
Although turnout was generally higher in 2016 than in 2018 (reflecting statewide higher turnout thanks to the presidential contest), there is no evidence that individuals discharged in the summer of 2016 were more likely to cast a ballot than individuals who finished parole earlier (the p-value on the coefficient of interest exceeds 0.5 in each model). The nonsignificant results from 2016 provides strong corroboration for causal claims regarding the efficacy of Executive Order 181 at boosting turnout.

>>>>

---
title: "Neighborhoods and Felony Disenfranchisement: The Case of New York City"
author: "Kevin Morris"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
bibliography: "../research/citations.bib"
link-citations: yes
---
<!-- --- -->
<!-- title: "Neighborhoods and Felony Disenfranchisement: The Case of New York City" -->
<!-- author: "Kevin Morris" -->
<!-- date: "`r format(Sys.time(), '%B %d, %Y')`" -->
<!-- output: word_document -->
<!-- bibliography: "../research/citations.bib" -->
<!-- link-citations: yes -->
<!-- always_allow_html: yes -->
<!-- --- -->

```{r setup, echo = F, message = F, warning = F}
knitr::opts_knit$set(root.dir = "..", dev = "cairo_pdf")
library(systemfit)
library(AER)
library(sandwich)
library(lmtest)
library(scales)
library(tidyverse)
library(extrafont)
library(kableExtra)
library(stargazer)
loadfonts()
```
# Introduction {-}

@@

## Identifying Parolees Whose Rights Were Restored {-}
Following Executive Order 181, the Department of Corrections and Community Supervision began indicating on their online Parolee Lookup Tool whether a parolee had her voting rights restored.^[Only individuals sentenced to prison for felony offenses lose their voting rights in New York. As such, I remove individuals sentenced to prison for misdemeanors. Incarceration likely also has a *de facto* disenfranchising effect on turnout among individuals incarcerated for misdemeanors [@White2019], though exploring that effect is beyond the scope of this project.] By using the DIN number provided from the parolee public records request, I was able to identify individuals who had their voting rights restored.^[Not all parolees listed in the public records request data are included in the lookup tool. For individuals who finished parole between January 1^st^, 2018, and April 17^th^, 2018, 1.0 percent are not in the lookup tool. For those discharged from parole between April 18^th^, 2018, and January 13^th^, 2019 (the latest date of the parole records), 1.2 percent of individuals are not found in the lookup tool.]

## Trends in Turnout {-}
Before analyzing turnout in the 2018 midterms, I begin by examining turnout in the 2016 election. It is possible that individuals discharged from parole shortly before a federal election are more likely to cast a ballot than individuals discharged earlier, whether or not their voting rights were restored. However, as Figure \ref{fig:to-16} makes clear, individuals discharged from parole in the final months before the 2016 presidential election were not substantially more likely to cast a ballot in the election than individuals discharged earlier in the year. The longer that individuals have been off of parole, the more likely they are to cast a ballot. For instance, of the individuals last discharged from parole in 2010, `r percent(readRDS("./temp/turnout_in_16_finished_10.rds"), accuracy = 0.1)` cast a ballot in the 2016 election, while just `r percent(readRDS("./temp/turnout_in_16_finished_15.rds"), accuracy = 0.1)` of those last discharged from parole in 2015 did so.^[Figure \ref{fig:to-16} plots individuals' turnout by the last date of discharge from parole. Therefore, individuals discharged from parole in 2010 who reoffended and were discharged from parole again in 2015 are included in the denominator only in 2015.] Figure \ref{fig:to-16} plots turnout rates by month of parole discharge. A quadratic curve is fitted (weighted by the number of individuals discharged each month), along with a 95 percent confidence band. This curve is fit on monthly data running from January, 2010 through April, 2016, and extended through October, 2016.

```{r to-16-chart, echo = FALSE, message = F, warning = F, fig.width = 7.5, fig.height = 4, fig.cap = "\\label{fig:to-16}Turnout in 2016 Presidential Election", fig.pos = "H", fig.align = "center"}
j <- readRDS("./temp/to16_chart.rds")
print(j)
```
Figure \ref{fig:to-18} plots month of parole discharge and turnout in the 2018 midterm elections. Once again, a weighted quadratic curve is fitted with a 95 percent confidence band. This curve is fit on monthly data running from January, 2012 through April, 2018, and extended through October, 2018.
```{r to-18-chart, echo = FALSE, message = F, warning = F, fig.width = 7.5, fig.height = 4, fig.cap = "\\label{fig:to-18}Turnout in 2018 Midterm Election", fig.pos = "H", fig.align = "center"}
j <- readRDS("./temp/to18_chart.rds")
print(j)
```
Figure \ref{fig:to-16} does not indicate that individuals who were discharged from parole shortly before the 2016 presidential election were more likely to cast a ballot than individuals discharged earlier in the year. Figure \ref{fig:to-18}, on the other hand, indicates that New Yorkers discharged from parole in the months leading up to the 2018 election --- many of whom had their rights restored while they were still on parole --- were more likely to participate than those discharged earlier in the year. However, Figures \ref{fig:to-16} and \ref{fig:to-18} are noisy. In the section that follows, I develop a logit model to explore whether, after controlling for the available characteristics, individuals whose rights were restored prior to being discharged were more likely to cast a ballot in the 2018 midterm elections.

@@

Figure \ref{fig:to-18} displays the share of individuals who were discharged from parole in each month and cast a ballot in the 2018 elections. It does not control for any sort of individual level characterisitcs. In \ref{tab:to-18-logit}, I present the results of an individual-level regression exploring the turnout in 2018 among individuals whose rights were restored. Because turnout is a binary dependent variable, this is a logistic regression. It includes all individuals discharged from parole between January 1^st^, 2012, through October 10^th^, 2018 (the registration deadline in New York State).^[This of course ignores the individuals who were still on parole and cast a ballot in the 2018 election. However, for this exercise, I limit the analysis group to individuals who would have been eligible to vote on election day even if Governor Cuomo had not signed Executive Order 181.]

@@

stargazer(model1, model2, model3,
          header = F,
          type = "latex",
          covariate.labels = c("D(Rights Restored)", "Days Since Discharged",
                               "Days Since Discharged\\textsuperscript{2}", "D(Male)", "Age (Years)", "Counts in Most Recent Sentence",
                               "Time Spent on Parole (Years)"),
          dep.var.labels = c("Cast Ballot in 2018 Election"),
          title = "(\\#tab:to-18-logit) Individual-Level Logit Model",
          table.placement = "H",
          omit = c("county", "race", "felony"),
          add.lines=list(c("County Fixed Effects", "", "X", "X"),
                         c("Race / Ethnicity Fixed Effects", "", "X", "X"),
                         c("Felony Class Fixed Effects", "", "", "X")),
          table.layout = "-dm#-t-a-s-n")
```

Model 1 in Table \ref{tab:to-18-logit} formalizes the trend presented in Figure \ref{fig:to-18} by controlling only for whether an individual had his rights restored before being discharged from parole, and the number of days between the individual's discharge date and November 6^th^, 2018. Model 2 also controls for individual-level characteristics: sex, age on November 6^th^, 2018, county, and race. Model 3 adds sentence-specific information to Model 2: the number of counts in the individual's most recent sentence, the amount of time they spent on parole, and the class(es) of felony for which they were convicted. In each successive model, the AIC decreases substantially, indicating that including these controls are warranted. Table \ref{tab:to-18-logit} makes clear that formerly incarcerated men were far less likely to vote than formerly incarcerated women; that older formerly incarcerated individuals were more likely to cast a ballot; and individuals who spent longer on parole were more likely to participate in the midterm election.

Each model also indicates that individuals whose rights were restored prior to discharge were more likely to cast a ballot than those whose rights were not restored. Exponentiating the coefficients on D(Rights Restored) indicates that individuals whose rights were restored due to Executive Order 181 were between `r round((exp(model1[["coefficients"]][["restoredTRUE"]]) - 1) * 100, digits = 1)` and `r round((exp(model3[["coefficients"]][["restoredTRUE"]]) - 1) * 100, digits = 1)` percent more likely to vote than those who did not.

Table \ref{tab:to-18-logit} provides some indication that Executive Order 181 was successful, but does not establish causality. There is reason to believe that individuals who had their rights restored were more likely to vote even had they not been "treated" by EO 181. In the section that follows, I employ an instrumental variables approach to control for this possibility.

@@

In the previous section, I explored whether individuals who had their rights formally restored prior to parole discharge were more likely to vote in the 2018 midterm election. There is reason to believe that whether an individual received the treatment (had their rights restored) is correlated with their propensity to vote. For instance, parolees who were arrested while on parole did not have their rights restored; a recent arrest is also likely to impact turnout. Similarly, noncitizens did not have voting rights restored / granted; such citizenship status is also of course correlated with propensity to vote. Because the treatment was not randomly assigned, direct comparison of turnout rates among individuals who did and did not have their rights restored is not probitive to the effect of Executive Order 181. This calls for an instrumental variables approach.

The timing of Executive Order 181 serves as a useful instrument for formal rights restoration. Although \ref{fig:to-18} shows that there is a relationship between how long an individual has been off parole and their propensity to vote over the long run, this relationship is far less signficant in the short run. Figure \ref{fig:to-18-recent} shows that, for individuals discharged from parole in 2017 and 2018, there does not appear to be a relationship between time off parole and turnout rates.

@@

Formalizing this chart into an individual-level logit model demonstrates that time since discharge is not correlated with turnout in the short run. The models in Table \ref{tab:to-18-logit-short} include individuals last discharged from parole between January 1^st^, 2017, and April 17^th^, 2018.

@@

Adding controls for time in Table \ref{tab:to-18-logit-short} increases the AIC, indicating that the inclusion of these variables is not warranted. A Chi-squared test confirms that the model is not improved when controls for time are included.^[[Appendix C] provides further corroboration that being discharged from parole in the months before an election is uncorrelated with propensity to vote by exploring turnout rates in the 2016 presidential election.]

Although date of discharge from parole is not correlated with turnout in the short run, it is correlated with likelihood of being having one's rights restored while still on parole. Governor Cuomo signed Executive Order 181 in April of 2018, but an examination of the individuals whose rights were ultimately restored indicates that the program did not go into full effect until later in May. As Figure \ref{fig:restoration-chart} demonstrates, individuals who finished their parole supervision in early May did not have their voting rights restored before they finished their term of supervision. It is not until May 21^st^ that the majority of parolees had their voting rights restored prior to their discharge.
```{r restoration-chart-chunk, echo = FALSE, message = F, warning = F, fig.width = 7.5, fig.height = 4, fig.cap = "\\label{fig:restoration-chart}Share of Discharged Parolees Whose Voting Rights Were Restored Prior to Discharge", fig.pos = "H", fig.align = "center"}
j <- readRDS("./temp/restoration_plot.rds")
print(j)
```
There is no evidence that date of discharged influences propensity to vote through channels other than formal rights restoration in the short run. A dummy variable indicating whether an individual was discharged on or after May 21^st^ can therefore effectively be used to instrument rights restoration. Table \ref{tab:iv-models} shows the second-stage estimation results, estimated using OLS.^[Although the dependent variable remains dichotomous (as do both the independent variable of interest and the instrument), @Angrist2008 recommends using OLS in instrumental variables analyses even under such conditions.]

```{r iv-reg-chunk-short, echo = F, results = "asis", message = F}
load("./temp/iv_individual_turnout_18.rdata")

summ.fit1 <- summary(iv1, diagnostics = T)
summ.fit2 <- summary(iv2, diagnostics = T)
summ.fit3 <- summary(iv3, diagnostics = T)

stargazer(iv1, iv2, iv3,
          header = F,
          type = "latex",
          covariate.labels = c("D(Rights Restored)", 
                               "D(Male)", "Age (Years)", "Counts in Most Recent Sentence",
                               "Time Spent on Parole (Years)"),
          dep.var.labels = c("Cast Ballot in 2018 Election"),
          title = "(\\#tab:iv-models) Second-Stage Regression",
          table.placement = "H",
          omit = c("county", "race", "felony"),
          add.lines=list(c("County Fixed Effects", "", "X", "X"),
                         c("Race / Ethnicity Fixed Effects", "", "X", "X"),
                         c("Felony Class Fixed Effects", "", "", "X"),
                         c("\\hline \\\\[-1.8ex]"),
                         c(rownames(summ.fit1$diagnostics)[1],
                           round(summ.fit1$diagnostics[1, "p-value"], 2),
                           round(summ.fit2$diagnostics[1, "p-value"], 2),
                           round(summ.fit3$diagnostics[1, "p-value"], 2)),
                         c(rownames(summ.fit1$diagnostics)[2],
                           round(summ.fit1$diagnostics[2, "p-value"], 2),
                           round(summ.fit2$diagnostics[2, "p-value"], 2),
                           round(summ.fit3$diagnostics[2, "p-value"], 2))),
          omit.stat = "ser",
          table.layout = "-dm#-t-as-n",
          notes = c("Table reports p-values for Wu-Hausman and",
                    "Weak Instruments tests.",
                    "D(Discharged from Parole on or after May",
                    "21, 2018) instrument for D(Rights Restored)"))
```

Table \ref{tab:iv-models} indicates that using D(Discharged from Parole on or After May 21) is likely warranted. The p-values of the Wu-Hausman tests hover around 0.05, while the Weak Instruments test is highly significant in each model.

Model 3 in Table \ref{tab:iv-models} indicates that having one's rights restored caused an individual's propensity to vote to increase by `r round(summary(iv3)[["coefficients"]][2,1] * 100, digits = 2)` percentage points. Approximately `r round(readRDS("./temp/to_restored.rds") * 100, digits = 2)` percent of individuals who had their rights restored cast a ballot in 2018, indicating that Executive Order 181 increased the propensity to vote by around `r round(summary(iv3)[["coefficients"]][2,1] / (readRDS("./temp/to_restored.rds") - summary(iv3)[["coefficients"]][2,1]) * 100, digits = 1)` percent. This estimate is much lower than the estimates reported in Table \ref{tab:to-18-logit}, but that is not surprising; as discussed above, individuals whose rights were restored were more likely to vote whether or not they received restoration before discharge. The `r round(summary(iv3)[["coefficients"]][2,1] * 100, digits = 2)` percentage point increase reported in Table \ref{tab:iv-models} accounts only for the extent to which rights restoration prior to discharge from parole caused higher turnout.

@@

In the [Instrumenting Rights Restoration] section of this paper, I argue that being discharged from parole in the final months leading up to an election is uncorrelated with propensity to vote. Table \ref{tab:to-16-logit} demonstrates that individuals discharged between May 21^st^ -- October 14^th^, 2016, did not participate at different rates in the 2016 presidential election than other formerly incarcerated individuals. Table \ref{tab:to-16-logit} includes all individuals last discharged from parole between January 1^st^, 2015, and October 14^th^, 2016.
```{r logit-reg-chunk-16, echo = F, results = "asis", message = F}
load("./temp/individual_turnout_16.rdata")

stargazer(model1, model2, model3,
          header = F,
          type = "latex",
          covariate.labels = c("D(Discharged on or after May 21\\textsuperscript{st}, 2016)", "Days Since Discharged",
                               "Days Since Discharged\\textsuperscript{2}", "D(Male)", "Age (Years)", "Counts in Most Recent Sentence",
                               "Time Spent on Parole (Years)"),
          dep.var.labels = c("Cast Ballot in 2016 Election"),
          title = "(\\#tab:to-16-logit) Individual-Level Logit Model",
          table.placement = "H",
          omit = c("county", "race", "felony"),
          add.lines=list(c("County Fixed Effects", "", "X", "X"),
                         c("Race / Ethnicity Fixed Effects", "", "X", "X"),
                         c("Felony Class Fixed Effects", "", "", "X")),
          table.layout = "-dm#-t-a-s-n")
```
Although turnout was generally higher in 2016 than in 2018 (reflecting statewide higher turnout thanks to the presidential contest), there is no evidence that being discharged in the summer of 2016 was associated with an individual's propensity to cast a ballot (the p-value on the coefficient of interest exceeds 0.5 in each model). The nonsignificant results from 2016 provide strong corroboration that discharge date serves as an effective instrument for rights restoration.
